---
title: "Available data on Makay protected area"
author: "Florent Bédécarrats"
format: 
  html: 
    toc: true
    output-file: index.html
    embed-resources: true
    standalone: true
    code-fold: true
execute:
  warning: false
  error: false
editor: visual
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

## Introduction

This interactive workbook aims at providing an overview of data sources that are potentially relevant for the analysis of economic, social and ecological dynamics in the area concerned by the creation of the Makay protected area.

## Working environment

We use R and a series of packages dedicated to spatial data and analysis.

```{r load_packages, warning=FALSE}

if (!"geoarrow" %in% installed.packages()) {
remotes::install_github("paleolimbot/geoarrow")
}


library(aws.s3)
library(tidyverse) # toolkit for data manipulation
library(geodata) # to get municipalities
library(sf) # to handle spatial data (vectors)
library(terra) # to handle patial data (rasters)
library(mapme.biodiversity) # to compute spatial indicators
library(tmap) # nice maps
library(zoo) # time series
library(units) # ensures units are right
library(future) # to parallelize computations
library(exactextractr) # engine for mapme.biodiversity
library(SPEI) # to compute rainfall  
library(geoarrow) # to write/read efficient spatial vector format
```

## Madagascar

National borders

```{r borders}
#| fig-cap: "Surrounding municipalities (source: GADM)"

# Download Madagascar municipalities
mada <- gadm("MDG", level  = 0, path = "data") %>%
  st_as_sf()  %>%
  st_cast("POLYGON")

```

## Forest cover and loss

This data relies on global forest cover for Madagascar

```{r map_forest}
#| fig-cap: "Forest cover in 2000 (source: GFW)"
# Reference portfolio (chirps data was loaded at startup)
mada <- init_portfolio(mada, years = 1990:2020,
                      outdir = "data", add_resources = FALSE) 
mada <- mada %>%
  get_resources(c("gfw_treecover", "gfw_lossyear"))

tifs_tc <- list.files(path = "data/gfw_treecover/", pattern = "tif", 
                      full.names = TRUE) 
tc <- sprc(lapply(tifs_tc, rast)) %>%
  mosaic()

tm_shape(tc) +
  tm_raster(title = "Forest cover in 2000 (%)", palette = "YlGn", style = "cont") +
  tm_shape(mada) +
  tm_layout(legend.outside = TRUE) 
```

Forest cover loss

```{r map_forest_loss}
#| fig-cap: "Forest cover in 2000 and loss since (source: GFW)"

tiffs_l <- list.files(path = "data/gfw_lossyear/", pattern = "tif", 
                     full.names = TRUE)

l <- sprc(lapply(tifs_l, rast)) %>%
  mosaic()

loss <- l
values(loss) <- ifelse(values(l) > 0, 1, NA)

tm_shape(tc_2000) +
  tm_raster(title = "Forest cover in 2000 (%)", palette = "YlGn", style = "cont") +
  tm_shape(loss) + 
  tm_raster(title = "Forest cover loss 2001-2021", 
            style = "cat", breaks = c(0, 0.5, 1.5), palette = "red") +
  tm_shape(mada)  

```
